```{r}
library(mice)
library(grpreg)
library(dplyr)

dataset <- read.csv("preprocessed_data.csv")
```

## Preprocess and Declare Variable Type

```{r}
# Declare variable types
categorical_vars <- c(
  "truant", "gor_x", "w2fat5yp", "w2ethnicyp",
  "w3outschbyp", "w3parqualmp", "w4concenyp", 
  "w4lnpar", "w4noconfyp", "w4readsyp",
  "w7heflag", "w9danvqh", "w9dvnvqh", "w9drussell",
  "marital_status_MP", "job_attd", "w2hiqualgmp"
)
binary_vars <- c("w6univyp", "w2sexyp", "drugs")
outcome <- "w9_income_above_median"

# convert to factors
dataset[categorical_vars] <- lapply(dataset[categorical_vars], as.factor)
dataset[binary_vars] <- lapply(dataset[binary_vars], as.factor)
dataset[[outcome]] <- as.factor(dataset[[outcome]])
```

```{r}
reset_baselines <- function(data, baselines) {
  for (var in names(baselines)) {
    if (var %in% names(data)) {
      data[[var]] <- relevel(factor(data[[var]]), ref = baselines[[var]])
    } else {
      warning(paste("Variable", var, "not found in data. Skipping."))
    }
  }
  return(data)
}

```


## Test Training Split
```{r}
set.seed(39748)
n <- nrow(dataset)
train_index <- sample(n, floor(0.8 * n))
training_data <- dataset[train_index, ]
testing_data  <- dataset[-train_index, ]
```

```{r}
# Define baselines
baselines <- list(
  gor_x = 7,
  w3parqualmp = 4,
  w4concenyp = 2,
  w4lnpar = 2,
  w9danvqh = 4,
  w9dvnvqh = 3,
  w9drussell = 3,
  marital_status_MP = 2,
  job_attd = 3,
  w2hiqualgmp = 3
)
```

```{r}
# Reset baseline levels
training_data <- reset_baselines(training_data, baselines)
testing_data  <- reset_baselines(testing_data, baselines)
```

## MICE Setup
```{r}
method <- rep("pmm", ncol(training_data))
names(method) <- colnames(training_data)
method[categorical_vars] <- "polyreg"
method[binary_vars] <- "logreg"
method[outcome] <- ""  

num_imputations <- 25
imp <- mice(training_data, m = num_imputations, method = method, seed = 39748, printFlag = TRUE)
```

## Multiple Imputations Diagnostics

```{r}
densityplot(imp, ~ w2fat5yp)
densityplot(imp, ~ w3parqualmp)
densityplot(imp, ~ w4ghq12scr)
densityplot(imp, ~ w4concenyp)
```

```{r}
plot(imp, c("w3parqualmp", "w2fat5yp"))
```
## Convergence Analysis
```{r}
library('miceadds')

psrf_df <- Rhat.mice(imp)
ac_df <- convergence(imp, diagnostic = "ac", parameter = "mean")
conv_all <- convergence(imp, diagnostic = "all", parameter = "mean")
head(conv_all)
```

## Apply Scaling to Continuous Predictors
```{r}
# Figure out which predictors are continuous (to scale)
# here numeric cols in training_data are the continuous vars
cont_vars <- names(training_data)[ sapply(training_data, is.numeric) ]
cont_vars <- setdiff(cont_vars, outcome)

# compute global mean & sd on original training data
gm <- colMeans(training_data[cont_vars], na.rm = TRUE)
gs <- apply(training_data[cont_vars], 2, sd, na.rm = TRUE)
```

## Setup for Model Fitting

```{r}
# Extract & scale each completed imputation
completed_list <- lapply(complete(imp, "all"), function(df) {
  # scale continuous
  df[cont_vars] <- scale(df[cont_vars], center = gm, scale = gs)
  df
})

# Prepare group indices (based on first completed df)
# build model.matrix once to get dummy‐expansion for factors
formula <- as.formula(paste(outcome, "~", paste(setdiff(names(training_data), outcome), collapse = "+")))
X0 <- model.matrix(formula, data = completed_list[[1]])[, -1]
colnames0 <- colnames(X0)

# For each original predictor, count how many columns it spawned
orig_preds <- setdiff(names(training_data), outcome)
group_ids <- rep(
  seq_along(orig_preds),
  times = sapply(orig_preds, function(v) sum(startsWith(colnames0, paste0(v))))
)
```


## Perform CV to find best lambda

```{r}
# Perform within-imputation CV to pick lambda
lambda_min <- numeric(num_imputations)

for (i in seq_len(num_imputations)) {
  df <- completed_list[[i]]
  X_i <- model.matrix(formula, data = df)[, -1]
  y_i <- as.numeric(df[[outcome]]) - 1
  
  cv_i <- cv.grpreg(
    X_i, y_i,
    group   = group_ids,
    penalty = "grLasso",
    family  = "binomial"
  )
  lambda_min[i] <- cv_i$lambda.min
}

# Pick a consensus lambda
lambda_star <- median(lambda_min)
lambda_star
```

## Identify Selected Predictors

```{r}
# Re-fit final models and tally selected predictors
selected_list <- vector("list", num_imputations)

for (i in seq_len(num_imputations)) {
  df  <- completed_list[[i]]
  X_i <- model.matrix(formula, data = df)[, -1]
  y_i <- as.numeric(df[[outcome]]) - 1

  fit_i <- grpreg(
    X_i, y_i,
    group   = group_ids,
    penalty = "grLasso",
    family  = "binomial",
    lambda  = lambda_star
  )
  coefs <- coef(fit_i)
  nz    <- which(coefs[-1] != 0)
  selected_list[[i]] <- unique(orig_preds[group_ids[nz]])
}

# Show selection frequency of predictors
sel_freq <- table(unlist(selected_list))

# Select those in >= 60% of imputations
final_vars <- names(sel_freq)[ sel_freq >= 0.6 * num_imputations ]

cat("Majority-selected predictors (≥60% of imputations):\n", paste(final_vars, collapse = ", "), "\n")
sel_freq
```

## Bar plot of Selected Predictors

```{r}
# Barplot of variable selection frequency
barplot(sel_freq[order(sel_freq, decreasing = TRUE)],
        las = 2, col = "skyblue",
        main = "Variable Selection Frequency",
        ylab = "Number of Imputations Selected")
abline(h = 0.6 * num_imputations, col = "red", lty = 2)
```

## Refit on Logistic Regression by Rubin's Rule

```{r}
# Group all majority predictors
final_form <- as.formula(
  paste(outcome, "~", paste(final_vars, collapse = "+"))
)

# Fit a logistic model on each imputation
fit_logit <- with(
  data = imp,
  expr = glm(final_form, family = binomial, data = training_data)
)

# Pool estimates
pooled   <- pool(fit_logit)
summary(pooled)
```

```{r}
library(pROC)

# storage
auc_vals <- numeric(num_imputations)
acc_vals <- numeric(num_imputations)

for(i in seq_len(num_imputations)){
  # 1. extract the i-th completed dataset
  dat_i <- complete(imp, action = i)

  # 2. fit the model on that completed data
  fit_i <- glm(final_form, family = binomial, data = dat_i)

  # 3. get predicted probabilities
  p_i <- predict(fit_i, type = "response")

  # 4. compute AUC
  roc_i    <- roc(dat_i[[outcome]], p_i)
  auc_vals[i] <- as.numeric(roc_i$auc)

  # 5. compute accuracy at threshold 0.5
  yhat_i  <- ifelse(p_i > 0.5, 1, 0)
  acc_vals[i] <- mean(yhat_i == dat_i[[outcome]])
}

# 6. pool by simple averaging
mean_auc <- mean(auc_vals)
mean_acc <- mean(acc_vals)
sd_auc <- sd(auc_vals)
sd_acc <- sd(acc_vals)

cat("Train set accuracy:", round(mean_acc, 4), "±", round(sd_acc, 4), "\n")
cat("Train set AUC:", round(mean_auc, 4), "±", round(sd_auc, 4), "\n")
```

## Compute Model Performance Metrics

```{r}
library(pROC)

# Impute & scale test set like training
imp_test <- mice(testing_data, m = num_imputations, method = method, seed = 456)
completed_test_list <- lapply(complete(imp_test, "all"), function(df) {
  df[cont_vars] <- scale(df[cont_vars], center = gm, scale = gs)
  df
})


# Extract unpenalised glm fits
fits <- fit_logit$analyses  

# Generate preds_list
preds_list <- lapply(
  seq_len(num_imputations),
  function(i) {
    predict(
      fits[[i]],
      newdata = completed_test_list[[i]],
      type    = "response"
    )
  }
)

# Store AUC and accuracy per imputation
auc_i <- numeric(num_imputations)
acc_i <- numeric(num_imputations)

for (i in seq_len(num_imputations)) {
  # Get predicted probabilities and true labels for this imputation
  pred_prob_i <- preds_list[[i]]
  actual_i    <- completed_test_list[[i]][[outcome]]

  # AUC
  roc_i <- roc(actual_i, pred_prob_i)
  auc_i[i] <- as.numeric(auc(roc_i))

  # Accuracy
  pred_class_i <- ifelse(pred_prob_i > 0.5,
                         levels(dataset[[outcome]])[2],
                         levels(dataset[[outcome]])[1])
  acc_i[i] <- mean(pred_class_i == actual_i)
}

# Report mean ± sd
cat("Test set accuracy:", round(mean(acc_i), 4), "±", round(sd(acc_i), 4), "\n")
cat("Test set AUC:", round(mean(auc_i), 4), "±", round(sd(auc_i), 4), "\n")
```
```{r}
library(caret)

# 1. Stack your preds_list into a matrix (rows = cases, cols = imputations)
preds_mat <- do.call(cbind, preds_list)   # dim: nrow(test) × num_imputations

# 2. Compute the mean predicted probability for each case
mean_pred_prob <- rowMeans(preds_mat)

# 3. Get the single “averaged” predicted class at cutoff = 0.5
avg_pred_class <- factor(
  ifelse(mean_pred_prob > 0.5,
         levels(dataset[[outcome]])[2],
         levels(dataset[[outcome]])[1]),
  levels = levels(dataset[[outcome]])
)

# 4. Pull out the true labels once (they’re the same across imputations)
actual_all <- completed_test_list[[1]][[outcome]]

# 5. Build the confusion matrix
cm_avg <- confusionMatrix(avg_pred_class, actual_all,
                          positive = levels(actual_all)[2])

# 6. View the raw 2×2 table and key metrics
cm_avg$table
cm_avg$overall["Accuracy"]
cm_avg$byClass[c("Sensitivity","Specificity")]
```