---
title: "XGBoost with Multiple Imputations"
author: "Joshua Ang"
date: "2025-04-29"
output: html_document
---

```{r}
library(dplyr)
library(ggplot2)
library(mgcv)
library(dplyr)
library(tidymodels)
library(rsample) 
library(pROC)
library(glmnet)
library(caret)
library(mice)
library(miceadds)
library(glmnet)
library(purrr)
library(rBayesianOptimization)
library(xgboost)
library(viridis)
library(SHAPforxgboost)
library(data.table)
library(ggnewscale)
```

```{r}
rm(list = ls())
```

```{r}
dataset <- read.csv('preprocessed_data.csv')
```

```{r}
dataset
```
## Convert Categorical and Binary Variables Into Factors

```{r}
categorical_vars <- c(
  "truant", "gor_x", "w2fat5yp", "w2ethnicyp",
  "w3outschbyp", "w3parqualmp", "w4concenyp", 
  "w4lnpar", "w4noconfyp", "w4readsyp",
  "w7heflag", "w9danvqh", "w9dvnvqh", "w9drussell", "marital_status_MP", 
  "job_attd", "w2hiqualgmp"
)

binary_vars <- c(
  "w6univyp", "w2sexyp", "drugs"
)

dataset[categorical_vars] <- lapply(dataset[categorical_vars], as.factor)
dataset[binary_vars] <- lapply(dataset[binary_vars], as.factor)
dataset$w9_income_above_median <- as.factor(dataset$w9_income_above_median)
```

## MICE Percentage Missing Check

```{r}
dataset_clean <- dataset %>% drop_na()

# Compute missing percentage
missing <- (nrow(dataset) - nrow(dataset_clean))/nrow(dataset)

# Use rule of n imputations for n% missing data
# n_imputations <- round(missing*100)
n_imputations <- 25
```

## Preprocess by Variable Type
```{r}
dataset[categorical_vars] <- lapply(dataset[categorical_vars], as.factor)
dataset[binary_vars] <- lapply(dataset[binary_vars], as.factor)
dataset$w9_income_above_median <- as.factor(dataset$w9_income_above_median)
```

```{r}
reset_baselines <- function(data, baselines) {
  for (var in names(baselines)) {
    if (var %in% names(data)) {
      data[[var]] <- relevel(factor(data[[var]]), ref = baselines[[var]])
    } else {
      warning(paste("Variable", var, "not found in data. Skipping."))
    }
  }
  return(data)
}
```

## Test Training Split
```{r}
set.seed(39748)
n <- nrow(dataset)
train_index  <- sample(n, size = floor(0.8 * n))
training_data <- dataset[ train_index, ]
testing_data  <- dataset[-train_index, ]
```

```{r}
# Define baselines
baselines <- list(
  gor_x = 7,
  w3parqualmp = 4,
  w4concenyp = 2,
  w4lnpar = 2,
  w9danvqh = 4,
  w9dvnvqh = 3,
  w9drussell = 3,
  marital_status_MP = 2,
  job_attd = 3,
  w2hiqualgmp = 3
)
```

```{r}
# Reset baseline levels
training_data <- reset_baselines(training_data, baselines)
testing_data  <- reset_baselines(testing_data, baselines)
```

## Setup MICE Vector
```{r}
# Use predictive mean matching (pmm) for continuous variables
# Treatment for continuous variables
method <- rep("pmm", ncol(dataset)) 
names(method) <- colnames(dataset)

# Treatment for categorical variables
method[categorical_vars] <- "polyreg"

# Treatement for binary  variables
method[binary_vars] <- "logreg"

# We do not perform any imputation on the target
method["w9_income_above_median"] <- ""
```

## Setup objective function for bayesian hyperparam search
```{r}
## master feature template: ensures consistent feature set
all_names <- colnames(
  model.matrix(w9_income_above_median ~ . -1, data = training_data)
)

## helper function: creates a DMatrix ensuring correct column alignment
build_dtrain <- function(data) {
  # Convert dataset to a model matrix
  X <- model.matrix(w9_income_above_median ~ . -1, data = data)

  # Identify missing columns compared to master template
  miss <- setdiff(all_names, colnames(X))
  
  # If columns are missing, add them filled with zeros
  if (length(miss)) {
    X <- cbind(X, matrix(0, nrow(X), length(miss), dimnames = list(NULL, miss)))
  }

  # Reorder columns to match master template exactly
  X <- X[, all_names, drop = FALSE]

  # Create binary label
  y <- as.integer(data$w9_income_above_median) - 1

  # Return as XGBoost DMatrix
  xgb.DMatrix(X, label = y)
}

## Generate a list of DMatrix objects, each based on a different imputation
dtrain_list <- lapply(seq_len(n_imputations), function(i) {
  # Impute missing data using MICE
  tra <- complete(mice(training_data, m = 1,
                       method = method,
                       seed = 2000 + i,
                       printFlag = TRUE))
  # Build the corresponding DMatrix
  build_dtrain(tra)
})

## Function generator: Creates an XGBoost cross-validation function used for Bayesian optimisation
make_xgb_cv_bayes <- function(dtrain_list, logfile) {

  function(max_depth, eta, subsample, colsample_bytree, min_child_weight) {

    tryCatch({
      # Sanitize hyperparameters to acceptable XGBoost ranges
      int_md  <- max(3L, round(max_depth))
      int_mcw <- max(1L, round(min_child_weight))
      eta <- pmin(pmax(eta, 1e-3), 1)
      subs <- pmin(pmax(subsample, 0.10), 1)
      csbt <- pmin(pmax(colsample_bytree, 0.10), 1)

      # Log current hyperparameters
      cat(sprintf("%s | PARAMS: max_depth=%d, eta=%.4f, subsample=%.2f, ",
                  Sys.time(), int_md, eta, subs),
          sprintf("colsample=%.2f, min_child=%d\n", csbt, int_mcw),
          file = logfile, append = TRUE)

      aucs  <- numeric(length(dtrain_list))
      iters <- integer(length(dtrain_list))

      # Perform cross-validation for each imputation
      for (i in seq_along(dtrain_list)) {

        cat(sprintf("  - [%s] Imputation %d start\n", Sys.time(), i),
            file = logfile, append = TRUE)

        res <- tryCatch({
          # Perform cross-validation with early stopping
          cv <- xgb.cv(
            params = list(
              objective = "binary:logistic",
              eval_metric = "auc",
              max_depth = int_md,
              eta = eta,
              subsample = subs,
              colsample_bytree = csbt,
              min_child_weight = int_mcw
            ),
            data = dtrain_list[[i]],
            nrounds = 500,
            nfold = 5,
            early_stopping_rounds = 20,
            verbose = FALSE
          )

          # Record best AUC and iteration
          auc_val  <- max(cv$evaluation_log$test_auc_mean)
          iter_val <- cv$best_iteration

          cat(sprintf("    --> AUC=%.5f, BestIter=%d\n", auc_val, iter_val),
              file = logfile, append = TRUE)

          list(auc = auc_val, iter = iter_val)

        }, error = function(e) {
          # Log any error during individual imputation runs
          cat(sprintf("    !! ERROR on imputation %d: %s\n", i, e$message),
              file = logfile, append = TRUE)
          list(auc = NA_real_, iter = NA_integer_)
        })

        cat(sprintf("  - [%s] Imputation %d end\n", Sys.time(), i),
            file = logfile, append = TRUE)

        aucs[i]  <- res$auc
        iters[i] <- res$iter
      } # end loop over imputations

      # Check if any finite results were obtained
      if (!any(is.finite(aucs)) || !any(is.finite(iters))) {
        cat("  !! No finite results, returning (0,0)\n\n",
            file = logfile, append = TRUE)
        return(list(Score = 0, Pred = 0))
      }

      # Compute average score and iteration across imputations
      final_score <- mean(aucs,  na.rm = TRUE)
      final_iter  <- round(mean(iters, na.rm = TRUE))

      cat(sprintf("  >> Returning: Score=%.5f, Pred=%d\n\n",
                  final_score, final_iter),
          file = logfile, append = TRUE)

      list(Score = final_score, Pred = final_iter)

    }, error = function(e) {
      # Handle top-level errors gracefully
      cat(sprintf("%s | !! TOP-LEVEL ERROR: %s\n", Sys.time(), e$message),
          file = logfile, append = TRUE)
      list(Score = 0, Pred = 0)
    })
  }  # end inner function
}

# Create log file name based on current timestamp
logfile <- paste0("bayes_opt_log_", format(Sys.time(), "%Y%m%d_%H%M"), ".txt")

# Instantiate Bayesian-optimized cross-validation function
xgb_cv_bayes <- make_xgb_cv_bayes(dtrain_list, logfile)

# Wrap safely to handle potential failures in optimisation
safe_xgb_cv <- possibly(
  .f = xgb_cv_bayes,
  otherwise = list(Score = 0, Pred = 0)
)
```

## Run Bayesian Optimisation

```{r}
# Set up hyperparam search space
bounds <- list(
  max_depth = c(3L, 10L),
  eta = c(0.001, 0.3),
  subsample = c(0.7, 1.0),
  colsample_bytree = c(0.7, 1.0),
  min_child_weight = c(3L, 15L)
)

set.seed(39748)

# Run optimisation with the objective function defined 
opt_res <- BayesianOptimization(
  FUN = safe_xgb_cv,
  bounds = bounds,
  init_points = 10,
  n_iter = 50,
  acq = "ei",
  verbose = TRUE
)
```

```{r}
## Extract best hyperparams across all hyperparam combos
best_pars <- opt_res$Best_Par
best_params <- list(
  objective = "binary:logistic",
  eval_metric = "auc",
  max_depth = as.integer(best_pars["max_depth"]),
  eta = best_pars["eta"],
  subsample = best_pars["subsample"],
  colsample_bytree = best_pars["colsample_bytree"],
  min_child_weight = as.integer(best_pars["min_child_weight"])
)

# Define function to process the hyperparam search history stored
get_best_from_opt <- function(opt_res) {
  h <- opt_res$History
  i <- which.max(h$Value)
  list(
    Score = h$Value[i],
    Round = h$Round[i]
  )
}

# Best number of boosted iteration rounds across all hyperparam combos
best <- get_best_from_opt(opt_res)
best_nrounds <- best$Round
best_nrounds
```
```{r}
library(ggplot2)
# pull out the history
df <- opt_res$History

# make depth a factor so it’s treated discretely
df$max_depth <- factor(df$max_depth)

# boxplot + jitter
ggplot(df, aes(x = max_depth, y = Value)) +
  geom_boxplot(outlier.shape = NA, alpha = 0.6) +
  geom_jitter(width = 0.2, size = 1, alpha = 0.8) +
  labs(
    x = "Tree Depth (max_depth)",
    y = "Validation AUC",
    title = "Distribution of AUC Scores by Tree Depth"
  ) +
  theme_minimal()

```


```{r}
library(dplyr)
library(ggplot2)

# add iteration counter and running incumbent
history_inc <- opt_res$History %>% 
  mutate(iter        = row_number(),
         best_so_far = cummax(Value))

# iteration where the global best is first met
stop_iter <- history_inc %>% 
  slice_max(best_so_far, n = 1, with_ties = FALSE)

ggplot() +
  # all individual trial results
  geom_point(data = history_inc,
             aes(iter, Value),
             colour = "grey60", alpha = 0.6, size = 1.3) +
  # incumbent staircase
  geom_line(data = history_inc,
            aes(iter, best_so_far),
            colour = "navy", linewidth = 0.9) +
  # mark the first best
  geom_point(data = stop_iter,
             aes(iter, best_so_far),
             colour = "red", size = 3) +
  labs(title = "Bayesian optimisation – AUC over iterations",
       x     = "Iteration",
       y     = "AUC") +
  theme_minimal(base_size = 12)
```

## Fit Final Boosted Tree based on Hyperparams Determined

```{r}
# Extract master feature names from training data
all_names <- colnames(
  model.matrix(w9_income_above_median ~ . -1, data = training_data)
)

# Build design matrix and ensure columns align with master list
realign_mm <- function(df) {
  X <- model.matrix(w9_income_above_median ~ . -1, data = df)
  miss <- setdiff(all_names, colnames(X))
  if (length(miss)) {
    X <- cbind(
      X,
      matrix(0, nrow(X), length(miss), dimnames = list(NULL, miss))
    )
  }
  X[, all_names, drop = FALSE]
}

# Prepare list to store final xgboost models
final_models <- vector("list", n_imputations)

# Log the start of the final fitting process
cat(Sys.time(), "\n", file = logfile, append = TRUE)
cat("  Starting final ensemble fit...\n", file = logfile, append = TRUE)

# Extract & scale each completed imputation
imp <- mice(training_data, m = n_imputations, method = method, seed = 39748, printFlag = TRUE)
completed_list <- lapply(complete(imp, "all"), function(df) {
  df
})

final_models <- vector("list", length(completed_list))

for (i in seq_along(completed_list)) {
  tra <- completed_list[[i]]
  X   <- realign_mm(tra)
  y   <- as.integer(tra$w9_income_above_median) - 1
  
  final_models[[i]] <- xgb.train(
    params  = best_params,
    data    = xgb.DMatrix(X, label = y),
    nrounds = best_nrounds,
    verbose = 1
  )
}
```


## Training Metrics
```{r}
library(pROC)

# 0) Identify rows with a valid outcome
idx <- which(!is.na(completed_list[[1]]$w9_income_above_median))

# 1) True labels (only for those rows)
true_y_sub <- as.integer(completed_list[[1]]$w9_income_above_median[idx]) - 1

# 2) Get matrix of predicted probabilities (only on idx rows)
train_preds <- sapply(seq_along(final_models), function(i) {
  X_all <- realign_mm(completed_list[[i]])
  X_sub <- X_all[idx, , drop = FALSE]
  predict(final_models[[i]], xgb.DMatrix(X_sub))
})

# 3) Compute per-imputation metrics
accuracies <- apply(train_preds > 0.5, 2, function(pred_bin) {
  mean(pred_bin == true_y_sub)
})

aucs <- apply(train_preds, 2, function(prob) {
  as.numeric(auc(roc(true_y_sub, prob)))
})

# 4) Summarize and print
metrics_df <- data.frame(
  imputation = seq_along(accuracies),
  accuracy   = accuracies,
  auc        = aucs
)

# define auc_val before using it
auc_val <- mean(aucs)

cat("Final training accuracy:", mean(accuracies), "±", sd(accuracies), "\n")
cat("Final training AUC: ", auc_val, "±", sd(aucs), "\n")
```

```{r}
# Identify rows with a valid income label
idx <- which(!is.na(completed_list[[1]]$w9_income_above_median))

# Build train_preds on exactly those rows
train_preds <- sapply(seq_along(final_models), function(i) {
  X_all <- realign_mm(completed_list[[i]])
  X_sub <- X_all[idx, , drop = FALSE]              # <-- subset here
  predict(final_models[[i]], xgb.DMatrix(X_sub))
})

# Match true_y to the same rows
true_y_sub <- as.integer(completed_list[[1]]$w9_income_above_median[idx]) - 1

# Sanity check
stopifnot(nrow(train_preds) == length(true_y_sub))

# Now you can compute a confusion matrix without error
pred_bin <- train_preds > 0.5
cm_pooled <- table(
  Predicted = as.integer(pred_bin[,1]),
  Actual    = true_y_sub
)
dimnames(cm_pooled) <- list(Predicted = c("0","1"), Actual = c("0","1"))
print(cm_pooled)
```

```{r}
# Test Set Evaluation
test_preds <- sapply(1:n_imputations, function(i) {
  te <- complete(mice(testing_data, m = 1, method = method, seed = 3000 + i, printFlag = FALSE))
  Xtest <- realign_mm(te)
  predict(final_models[[i]], xgb.DMatrix(Xtest))
})

final_prob  <- rowMeans(test_preds)
final_prob_sd <- apply(test_preds, 1, sd)
final_class <- ifelse(final_prob > 0.5, 1, 0)

true_y <- as.integer(testing_data$w9_income_above_median) - 1

accuracy_i <- sapply(1:n_imputations, function(i) {
  pred_class_i <- ifelse(test_preds[, i] > 0.5, 1, 0)
  mean(pred_class_i == true_y)
})
auc_i <- sapply(1:n_imputations, function(i) {
  auc( roc(true_y, test_preds[, i]) )
})

cat("Final test accuracy:", mean(accuracy_i),"±", sd(accuracy_i), "\n")
cat("Final test AUC: ", mean(auc_i),"±", sd(auc_i), "\n")
```

## Visualise Variable Importance

```{r}
# Extract variable importance from each model
importance_list <- lapply(final_models, function(model) {
  imp <- xgb.importance(model = model)[, c("Feature", "Gain")]
  colnames(imp) <- c("Feature", "Gain")
  imp
})

# Combine into one data frame
importance_df <- bind_rows(importance_list, .id = "Imputation")

# Compute mean Gain and standard error per feature
importance_summary <- importance_df %>%
  group_by(Feature) %>%
  summarise(
    GainMean = mean(Gain, na.rm = TRUE),
    Variability = sd(Gain, na.rm = TRUE)
  ) %>%
  arrange(desc(GainMean))

# Plot top 10 features by mean Gain, with standard error bars
top_n <- 10
importance_plot <- importance_summary[1:top_n, ]

ggplot(importance_plot, aes(x = reorder(Feature, GainMean), y = GainMean)) +
  geom_col(fill = "skyblue") +
  geom_errorbar(aes(ymin = GainMean - Variability, ymax = GainMean + Variability),
                width = 0.2, color = "black") +
  coord_flip() +
  labs(title = "Variable Importance with Standard Error (Top 20)",
       x = "Feature",
       y = "Mean Gain") +
  theme_minimal()
```

## Compute SHAP Values

```{r}
# Impute once, build X_test
test_imputed <- complete(
  mice(testing_data, m = 1, method = method, seed = 39748, printFlag = TRUE)
)
X_test <- realign_mm(test_imputed)

# Compute per-model SHAP
shap_raw_list  <- list()
mean_shap_list <- list()

for (i in seq_along(final_models)) {
  sv <- shap.values(xgb_model = final_models[[i]], X_train = X_test)
  shap_raw_list[[i]]  <- as.matrix(sv$shap_score)
  mean_shap_list[[i]] <- sv$mean_shap_score
}

# Average across imputations
shap_avg        <- Reduce(`+`, shap_raw_list)  / length(shap_raw_list)
mean_shap_score <- Reduce(`+`, mean_shap_list) / length(mean_shap_list)

# Top features by mean
top10_features <- names(sort(mean_shap_score, decreasing = TRUE)[1:10])

shap_long_top10 %>% filter(!variable %in% exclude_vars)

shap_long <- shap.prep(
  shap_contrib = as.data.frame(shap_avg),
  X_train      = X_test
) %>%
  filter(variable %in% top10_features, ! variable %in% exclude_vars) %>%
  mutate(variable = factor(variable, levels = rev(top10_features)))

# Identify binary variables based on unique raw values
bin_vars <- shap_long %>%
  group_by(variable) %>%
  summarise(nval = n_distinct(rfvalue, na.rm = TRUE), .groups = "drop") %>%
  filter(nval == 2) %>%
  pull(variable)

# Split data into binary and continuous
shap_bin  <- filter(shap_long_top10, variable %in% bin_vars)

# Plot
ggplot() +
  new_scale_colour() +
  geom_point(
    data = shap_bin,
    aes(x = value, y = variable, colour = factor(rfvalue, levels = c(0, 1))),
    alpha = 0.8, size = 1.8,
    position = position_jitter(width = 0, height = 0.2)
  ) +
  scale_colour_manual(
    name = "Raw value (binary)",
    values = c("0" = "black", "1" = "blue"),
    drop = FALSE
  ) +
  
  labs(
    title = "SHAP summary – Top 10 features",
    x = "SHAP value",
    y = NULL
  ) +
  theme_minimal(base_size = 11) +
  theme(
    axis.text.y = element_text(size = 10),
    axis.text.x = element_text(size = 9),
    legend.title = element_text(size = 9),
    legend.text = element_text(size = 8)
  )
```

## Explore Probability Delta due to Feature

```{r}
# Grab each model’s intercept
bias0_list <- numeric(length(final_models))
for (i in seq_along(final_models)) {
  sv <- shap.values(
    xgb_model = final_models[[i]],
    X_train   = X_test
  )
  bias0_list[i] <- as.numeric(sv$BIAS0)
}

# Average the intercepts to get a single baseline og-odds
mean_bias0 <- mean(bias0_list)

# Build a data‐frame of mean SHAP + probability deltas
library(tibble)

prob_delta_df <- tibble(
  variable  = names(mean_shap_score),
  mean_shap = unname(mean_shap_score),
  p_before  = plogis(mean_bias0),
  p_after   = plogis(mean_bias0 + mean_shap),
  delta_p   = p_after - p_before
) %>%
  arrange(desc(delta_p))

# Add a log-ratio column (natural log)
prob_delta_df <- prob_delta_df %>%
  mutate(
    log_ratio = log(p_after / p_before)
  )

# Top 10 features:
top10_mean_delta <- prob_delta_df[order(-prob_delta_df$delta_p), ][1:10, ]
print(top10_mean_delta)
```
## Explore Interactions

```{r}
# Randomly select a model to interpret
selected_model <- sample(final_models, 1)[[1]]

# Compute usual SHAP values in long-format table
shap_res   <- shap.values(xgb_model = selected_model, X_train = X_test)
shap_long  <- shap.prep(shap_contrib = shap_res$shap_score, X_train = X_test)

# Compute the full SHAP interaction array
# This returns an [n_obs x n_feat x n_feat] array of pairwise interaction contributions
shap_int <- shap.prep.interaction(
  xgb_model = selected_model,
  X_train   = X_test
)  # :contentReference[oaicite:0]{index=0}

# Compute the mean(|interaction|) for each feature‐pair
# shap_int is an array [n_obs × n_feat × n_feat]
int_strength <- apply(abs(shap_int), c(2, 3), mean)

# Turn it into a long tibble for ggplot
int_df <- as.data.frame(int_strength) %>%
  rownames_to_column("Feature1") %>%
  pivot_longer(-Feature1, names_to="Feature2", values_to="MeanAbsInteraction")

# Filter for only unique pairs & threshold > 0.01
int_df_sig <- int_df %>%
  filter(Feature1 != Feature2,
         MeanAbsInteraction > 0.01) %>%
  arrange(desc(MeanAbsInteraction))
int_df_sig
```
